---
Name: Reginald Mordi
Course: Machine Learning 2020
Date: 10/12/2020
output:
  html_document:
    df_print: paged
---
```{r}
library(dplyr)
```


```{r}
NB <- read.csv("FlightDelays.csv")

# converting week and Time Variables to factors
# converting the target variable from character to factor
NB$CRS_DEP_TIME<-as.factor(NB$CRS_DEP_TIME)
NB$DAY_WEEK<-as.factor(NB$DAY_WEEK)
NB$Flight.Status<-as.factor(NB$Flight.Status)

data <- NB[,c(1, 2, 4, 8, 10, 13)]
str(NB)

```

1. Divide the data into 60% training and 40% validation
```{r}
library(caret)
set.seed(123)
Index_Train<-createDataPartition(data$Flight.Status, p=0.6, list=FALSE) # Use 60% of data for training and the rest for validation
Train <-data[Index_Train,]
Valid <-data[-Index_Train,]
```

2. Run the Naive Bayes model to predict whether the flight is delayed or not. Use only categorical variables for the predictor variables. Note that Week and Time variables need to recoded as factors

```{r}
# Build a na?ve Bayes classifier
library(e1071)
NB_model <-naiveBayes(Flight.Status~CRS_DEP_TIME+CARRIER+DEST+ORIGIN+DAY_WEEK,data = Train)
NB_model
```

3.	Output both a counts table and a proportion table outlining how many and what proportion of flights were delayed and on-time at each of the three airports.

```{r}
library(descr)
CrossTable(x=Train$DEST,y=Train$Flight.Status, prop.chisq = FALSE) 

prop.table(table(Train$DEST, Train$Flight.Status))
prop.table(table(Train$DEST, Train$Flight.Status),1)
prop.table(table(Train$DEST, Train$Flight.Status),2)
```

4. Output the confusion matrix and ROC for the validation data

```{r}
library(ISLR)
# Predict the Flight status of Valid dataset 
Predicted_Valid_labels <-predict(NB_model,Valid)
CrossTable(x=Valid$Flight.Status,y=Predicted_Valid_labels, prop.chisq = FALSE) 

# Show the confusion matrix of the classifier
confusionMatrix(Predicted_Valid_labels, Valid$Flight.Status)

# Make predictions and return probability of each class
Predicted_V_labels <-predict(NB_model,Valid, type = "raw")
# show the first few values 
head(Predicted_V_labels)

library(pROC)
# Passing the second column of the predicted probabilities 
# That column contains the probability associate to 'ontime'
roc(Valid$Flight.Status, Predicted_V_labels[,2])

plot.roc(Valid$Flight.Status,Predicted_V_labels[,2])
```